---
title: "Malaria_cases"
author: "Mei Qi Ng"
format: html
editor: visual
execute:
  echo: true
---

## Approach

For this assignment, I am using the Malaria dataset derived from the csv files downloaded from Kaggle. This data set provides information about the estimated number of malaria cases across the world.

Dataset Source: https://raw.githubusercontent.com/meiqing39/DATA-607/refs/heads/main/Malaria_estimated_numbers.csv

Assigned name and loaded the raw URL file of Malaria dataset from Github. Also using tidyverse

```{r}
malaria <- read.csv("https://raw.githubusercontent.com/meiqing39/DATA-607/refs/heads/main/Malaria_estimated_numbers.csv")

library(tidyverse)
```

All codes used here were from R for Data Science(2e) book

#Code Base #Checking to see if I can open data as a tibble data frame and first few columns

```{r}
glimpse(malaria)

```

#Select, rename, drop missing values on data set. assign another name in order to be used for creation of scatter plot

I used dpylr's select(),rename(), then drop_na to clean After inspecting the data, I see there are missing data and the column names are difficult to use since there are many period. I selected the few columns and rename them. I removed the ones that had NA or no values in the data subsets.

```{r}
malaria_clean <- malaria |> select(Country,Year,No..of.cases_max, No..of.deaths_max) |> rename(country = Country, year = Year, max_cases = No..of.cases_max, max_deaths = No..of.deaths_max)|> drop_na(max_cases, max_deaths)
```

#Max cases vs Max death cases scatterplot

Using tidyverse ggplot2() to explore the relation between max number of malaria cases vs the max number of deaths by malaria per country

```{r}
ggplot(malaria_clean, aes(x=max_cases, y=max_deaths))+ geom_point()
```

#Conclusion

In this assigment, I loaded dataset from Github, inspected the structure, cleaned it using select(), rename(), drop_na(), and created a scatterplot to explore the relationships in the data. There is a clear positive relatioship between max malaria cases and deaths. To further analyze this, I could analyze trends across years, compare different countries , and check the dataset with WHO source for values accuracies. I can also add more variables and other data visualization to present more information and deepen the analysis.
