---
title: "Assignment_3A Approach"
author: "Mei Qi Ng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Approach
The Goal of this assignment is to use a Global Baseline Estimate (GBE) recommendation system using R and see how personalization changes results in predictions. With the movie rating data collected in the previous assignment (Assignment_2A_Movie ratings), I aim to predict rating values for movies that reviewers have not yet seen. This is possible with the GBE formula. The dataset has missing values (NA/NULL) and these will be the target for our prediction algorithm. This model will account for reviewers movie bias preferences and popularity towards the item bias. 

Dataset contains ratings of 6 reviewers (Liz, Jed, Brenda, Jamie, Justice, Theresa) and 6 horror/thriller movies(The Substance, Nosferatu, Frankenstein, 28 years later, Sinners, The Conjuring 4) 

Source: Dataset from local PostgreSQL populated via Assigment2A_movie_rating.sql


# Implementation Steps 

(1) Connecting/Loading Data

        PostgreSQL database connection with DBI and RPostgres, import raw ratings table into R data frame, and convert NULL 
        values in SQL into NA values in R for calculation

(2) Bias Calculation and apply GBE forumla
     With dyplyr, use Group_by function and calculate: 
      - Mean of all rating 
      - Group: title --> movies bias
      - Group: reviewer_name --> User bias 
    GBE formula for User+movie pair 
(3) Comparison and Data Visualization

       Raw average vs GBE average predictions and seeing the user bias distribution. Find the predicted score for a reviewer, in this       case,   I will use Liz as an example. 

```{r}

```

