---
title: "Assignment NYTAPI Approach"
author: "Mei Qi Ng"
format: pdf
editor: visual
toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Approach

For this data acquisition assignment, my task will be to work with New York Times Books API, using the `/lists/current/hardcover-fiction.json` endpoint. This API shows the current top bestseller for hardcover fiction, along with other information such as author, publisher, and the number of weeks the book has been on the list

Question that I will attempt to answer using data analysis:

**What are the characteristics of the current top-ranking hardcover fiction books, and which publishers dominate the list currently?**

## Planned Approach steps

1.  **Secure Setup:** I will obtain an API key from the NYT Developer Network. To avoid hard coding in my script, I will store the key locally in an `.Renviron` file and retrieve it using `Sys.getenv()`as the environmental variable.This way this file will be reproducible and the system will look for local .Renviron file  of the users and their "NYT_API_KEY" from the New York Times Developer website. 

2.  **Client-Server Communication (The Request):** Rather than scraping unstructured HTML directly from the New York Times website, I will interact with the NYT server with the `httr` package. This allows me send an HTTP `GET` request to the API endpoint (`/lists/current/hardcover-fiction.json`)which will act as the exact address on the NYT server where the current best seller data is stored. I will also use `stop_for_status()` to catch any HTTP errors (e.g., a 404 Not Found) just in case.

3.  **Parsing the JSON Payload (The Response):** Once GET request completes successfully, the NYT server will return a response with the requested data formatted as JSON. JSON is the standard data format for web APIs because it organizes information into simple "key: value" pairs. I will use the `jsonlite` package and its `fromJSON()` function to translate this text file into a list that R can easily read

4.  **Data Tidying:** I will extract the target data (`books` array) and changing it into tidy tibble table with the dplyr function from R tidyverse. This process involves choosing specific columns needed to answer my question.

## Anticipated Data Challenges

While the NYT API is well documented, I believe the few challenges that I may run into during the data transformation phase:

-   **Nested Datas:** JSON files often contain lists inside of lists. One book might have a sublist of many different ISBN numbers or purchase links. When I bring this into R, these sublists can make the data frame hard to work with. I will need to select only the top level information (like the title and author) to keep my final table clean and flat.

-   **Checking Data Types:** When importing data from the web, numbers are sometimes accidentally read as text. I will need to check that columns like the book's `rank` or `weeks_on_list` are formatted as numbers in R so I can sort and analyze them properly.

## Source

-   Extracting Data from the Web Part 1: APIs (Lecture Notes by Garrett Grolemund)
-   Extracting Data from the Web Part 2: Web Scraping (Lecture Notes by Garrett Grolemund)
-   R for Data Science (2e): https://r4ds.hadley.nz/
-   New York Times Developer Network - Books API: https://developer.nytimes.com/docs/books-product/1/overview


## Code Base
```{r}
# Load required package libraries
library(httr)
library(jsonlite)
library(tidyverse) 

# 1. Secure Setup: Key retrieve from local .Renviron file

nyt_api <- Sys.getenv("NYT_API_KEY")

#Endpoint URL
url <- "https://api.nytimes.com/svc/books/v3/lists/current/hardcover-fiction.json"

# 2. Client-Server Communication

server_reply <- GET(url, query = list('api-key' = nyt_api))

stop_for_status(server_reply) #if HTTP error, stops script 

# 3. JSON Parsing

json_text <- content(server_reply, "text", encoding = "UTF-8") #Extract raw text from server

parsed_text <- fromJSON(json_text) #parsed raw json text into list

books_raw <- as_tibble(parsed_data$results$books)

# Flatten and clean the data to answer your analytical question
tidy_books <- books_raw %>%
  select(
    rank,
    title,
    author,
    publisher,
    weeks_on_list
  )

# Display the final, clean table
head(tidy_books)
```

