---
title: "Assignment 10B Nobel Prize API JSON Practice"
author: "Mei Qi Ng & Samantha Barbaro"
format: html
toc: true
toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
---

## Approach

The Nobel Prize organization provides public APIs for accessing structured Nobel Prize data.

In this assignment, we will be using the Nobel Prize **API v2.1** to retrieve JSON data format which includes information on the laureates, categories, award years, and birthplaces etc.

The data-driven questions we are investigating for this analysis are:

**1. Which Nobel Prize categories have the highest number of shared prizes (more than one laureate) vs which has the highest number of solo winners?**

**2. Which countries have the most Nobel laureates by birth ?**

**3. How has the number of Nobel Prizes awarded changed over time?**

**4. Which Nobel laureates received multiple Nobel Prizes, and in which categories?**

The data retrieved from the public API endpoints from Nobel Prize developer website will be transformed into tidy data frames for analysis (cleaned, joined, filtered, and compared) to answer the 4 questions by exploring the trends, patterns, and characteristics of the Nobel prize awards using R.

## Sources

-   Nobel Prize Outreach. (2026). *Nobel Prize Developer Zone – API Version 2.1*.\
    https://www.nobelprize.org/about/developer-zone-2/

-   Nobel Prize Outreach. (2026). *Nobel Prize API Documentation (v2.1)*.\
    https://api.nobelprize.org/2.1/

## Set up
```{r}
library(jsonlite)
library(tidyverse)
library(httr)


#Retrieve laureates JSON from API on Nobel Prize developer zone API

url_l <- "https://api.nobelprize.org/2.1/laureates"

response <- GET(url_l)
json_l <- content(response, as = "text", encoding = "UTF-8")

data_l <- fromJSON(json_l, flatten = TRUE)

laureates <- as_tibble(data_l$laureates)


#Retrieve Nobel Prize JSON from API on Nobel Prize developer zone API

prize_url <- "https://api.nobelprize.org/2.1/nobelPrizes"

response <- GET(prize_url)
prize_json <- content(response, as = "text", encoding = "UTF-8")

prize_data <- fromJSON(prize_json, flatten = TRUE)

n_prizes <- as_tibble(prize_data$nobelPrizes)


```



### Question 1: Which Nobel Prize categories have the highest number of shared prizes (more than one laureate) vs which has the highest number of solo winners?

```{r}
 shared_prizes <- n_prizes %>%
  unnest(nobelPrizes.laureates) %>%
  group_by(awardYear, category.en) %>%
  summarise(num_laureates = n(), .groups = "drop") %>%
  mutate(shared = num_laureates > 1)

shared_prizes %>%
  group_by(category.en) %>%
  summarise(
    total_prizes = n(),
    shared_prizes = sum(shared),
    share_rate = shared_prizes / total_prizes
  ) %>%
  arrange(desc(share_rate))

```
