---
title: "Assignment 10B Nobel Prize API JSON Practice"
author: "Mei Qi Ng & Samantha Barbaro"
format: html
toc: true
toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
---

## Approach

The Nobel Prize organization provides public APIs for accessing structured Nobel Prize data.

In this assignment, we will be using the Nobel Prize **API v2.1** to retrieve JSON data format which includes information on the laureates, categories, award years, and birthplaces etc.

The data-driven questions we are investigating for this analysis are:

**1. Which Nobel Prize categories have the highest number of shared prizes (more than one laureate) vs which has the highest number of solo winners?**

**2. Which countries have the most Nobel laureates by birth ?**

**3. How has the number of Nobel Prizes awarded changed over time?**

**4. Which Nobel laureates received multiple Nobel Prizes, and in which categories?**

The data retrieved from the public API endpoints from Nobel Prize developer website will be transformed into tidy data frames for analysis (cleaned, joined, filtered, and compared) to answer the 4 questions by exploring the trends, patterns, and characteristics of the Nobel prize awards using R.

## Sources

-   Nobel Prize Outreach. (2026). *Nobel Prize Developer Zone – API Version 2.1*.\
    https://www.nobelprize.org/about/developer-zone-2/

-   Nobel Prize Outreach. (2026). *Nobel Prize API Documentation (v2.1)*.\
    https://api.nobelprize.org/2.1/

## Set up
```{r}
library(jsonlite)
library(tidyverse)
library(httr)


#Retrieve laureates JSON from API on Nobel Prize developer zone API

url_l <- "https://api.nobelprize.org/2.1/laureates?limit=1000"  #add limit parameter for more data to find more reliable results

response <- GET(url_l)
json_l <- content(response, as = "text", encoding = "UTF-8")

data_l <- fromJSON(json_l, flatten = TRUE)

laureates_1<- as_tibble(data_l$laureates)


#Retrieve Nobel Prize JSON from API on Nobel Prize developer zone API

prize_url <- "https://api.nobelprize.org/2.1/nobelPrizes?limit=1000"

response <- GET(prize_url)
prize_json <- content(response, as = "text", encoding = "UTF-8")
prize_data <- fromJSON(prize_json, flatten = TRUE)
n_prizes <- as_tibble(prize_data$nobelPrizes)


```

## Question 1: Which Nobel Prize categories have the highest number of shared prizes (more than one laureate) vs which has the highest number of solo winners?

**The number of shared prizes was found by counting the number of laureates associated with each Nobel Prize. For prizes with more than 1 laureate, it was classified as shared while single laureate were known as solo. The classification was summarized using Nobel prize category to find out which was most frequently awarded shared solo prizes.Based on the table below, Economic Sciences has the highest number and proportion of shared prizes, 57 out of 57 prizes were shared but if we look at just numbers, then Physics has the highest number of shared prizes (119) total. As for the highest number of solo awards, that would be Peace category (19).**
```{r}
 shared_prizes <- n_prizes |> 
  mutate(num_laureates = map_int(laureates, length)) %>%  #Returns integers for each laureate list column
  mutate(shared = num_laureates > 1)  #Create shared prize, for number of laureates for prize greater than 1


shared_prizes |> 
  group_by(category.en) |> #group by the type of award categories
  summarise(
    total_prizes = n(),
    shared_prizes = sum(shared),
    solo_prizes = sum(!shared),
    share_rate = shared_prizes / total_prizes,
    .groups = "drop") |>
  arrange(desc(share_rate))

```

## Question 2: Which countries have the most Nobel laureates by birth ?

**To identify which countries have the most Nobel laureates by birth, the laureates data set was group by country of birth and number of laureates from each country was counted. The United states has the highest number of Nobel laureates by birth, followed by United kingdom, Germany, and France. From this result, we can see that these countries have historically been major centers for scientific, literacy, and political contributions recognized by the Nobel Prizes.**
```{r}

most_laureates <- laureates_1 |> 
  filter(!is.na(birth.place.countryNow.en)) |> #filter out NA
  group_by(birth.place.countryNow.en) |> 
  count() |> 
  arrange(desc(n))

```
## Question 3: How has the number of Nobel Prizes awarded changed over time?

**The number of Nobel Prizes awarded each year has remained consistent overtime. Five prizes were awarded annually from 1901 through 1968. In 1969, a structural increase to six prizes occurred with the introduction of the Nobel Memorial Prize in Economic Sciences, after which the annual number of prizes has remained stable. This indicates that changes over time are driven by institutional addition rather than gradual growth.**

```{r}
# Count number of awards over time
changed_nbp <- n_prizes |> 
  mutate(awardYear = as.integer(awardYear)) |>  #ensure variable from API is numeric
  count(awardYear, name = "num_prizes")

print(changed_nbp, nrow(5))

# Plot

ggplot(changed_nbp, aes(x=awardYear, y = num_prizes))+
  geom_line()+
  geom_point(size = 2)+ 
  geom_vline(xintercept = 1969, linetype = "dashed", color = "blue")+
  annotate("text", x = 1973, y = 5.5, label = "Economic Nobel Prize Introduced", color = "red")+  #breakaway annotation explanation
  scale_x_continuous(breaks = seq(1900, 2025, by = 10))+ # format x axis (shows every 10 years)
  labs(title = "Nobel Prizes Awarded Over Time",
       x = "Award Year",
       y = "Annual Prizes Awarded")

```
## Question 4: Which Nobel laureates received multiple Nobel Prizes, and in which categories? 


**By joining the Nobel Prize and laureate datasets, it is possible to identify individuals who received more than one Nobel Prize. Only a few laureates have won multiple awards, often across different categories, which shows the rarity of multiple Nobel recognition and impact of these individual's contributes.These laureates include Marie Curie, who won prizes in both Physics and Chemistry, and Linus Pauling, who received awards in Chemistry and Peace. Other repeat winners are John Bardeen, Frederick Sanger, and K. Barry Sharpless who all received multiple prizes within the same scientific category. **
```{r}

# create 1 row per laureate to prize combination

nbprize_laureates <- n_prizes |>
  select(awardYear, category.en, laureates) |>
  unnest(laureates) |>
  transmute(
    laureate_id = id,
    awardYear,
    category = category.en
  )


glimpse(nbprize_laureates)


#Join with laureates_1 table to get names


laureate_prizes <- nbprize_laureates |>
  left_join(
    laureates_1 |>
      transmute(
        laureate_id = id,
        full_name = fullName.en
      ),
    by = "laureate_id"
  )


# Find laureates with multiple prizes


multiple_winners <- laureate_prizes |>
  group_by(laureate_id, full_name) |>
  drop_na(full_name) |> 
  summarise(
    num_prizes = n(),
    categories = paste(unique(category), collapse = ", "),
    .groups = "drop"
  ) |>
  filter(num_prizes > 1) |>
  arrange(desc(num_prizes))

```
